Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2023-10-16 manu-draft.qmd
intro revision from rn comments

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method"
bibliography: refs.bib
csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Jill M Arriola:
      email: jva5648@psu.edu 
      institute: [PSU]
      correspondence: false
  - Maria Hermann:
      email: mxh367@psu.edu
      institute: [PSU]
      correspondence: false
  - Raymond G Najjar:
      email: rgn1@psu.edu
      institute: [PSU]
      correspondence: false

@@

Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism. Most models for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen data, which can lead to poorly constrained errors that influence the interpretation of results. In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems. This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, the Estuarine BAyesian Single-station Estimation (EBASE) method. EBASE provides several advantages over existing models by using an integration formula appropriate for estuarine systems and using a Bayesian framework where critical model parameters are fit as a function of the data as informed by a set of prior distributions. This approach can also vary the time period over which critical model parameters are estimated. We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola estuarine reserve and provide a quantitative assessment of how the prior distributions affect the model results. The approach is also bench-marked with more conventional methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration). The EBASE approach is available using open source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

@@

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecosystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011]. Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005]. However, gross primary production (P) and ecosystem respiration (R) estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are further compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of P and R. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole system respiration rates based on the open-water method [@hoellein2013].

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This approach exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (net primary production $+$ autotrophic respiration), ecosystem respiration (heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS). Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this challenge is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites. Estimating gas exchange using the open-water method in estuaries has also been challenging. Most approaches estimate gas exchange using empirical formula dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18]. Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimations based on empirical formulae or fixed values can lead to poorly constrained errors that influence interpretation of results.

A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Bayesian techniques have only been explored in limited estuarine applications despite their potential to address the above \[ @ciavatta2008; @tassone2019\]. Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016]. These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of parameters related to gas exchange, P, and R (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange. A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15]. BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations. As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism, and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

>>>>

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)"
bibliography: refs.bib
csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Jill M Arriola:
      email: jva5648@psu.edu 
      institute: [PSU]
      correspondence: false
  - Maria Hermann:
      email: mxh367@psu.edu
      institute: [PSU]
      correspondence: false
  - Raymond G Najjar:
      email: rgn1@psu.edu
      institute: [PSU]
      correspondence: false

@@

Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism. Most methods for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen data, which can lead to poorly constrained errors that influence the interpretation of results. In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems. This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, the Estuarine BAyesian Single-station Estimation (EBASE) method. EBASE provides several advantages over existing models by using an integration formula appropriate for estuarine systems and using a Bayesian framework where critical model parameters are fit as a function of the data as informed by a set of prior distributions. This approach can also vary the time period over which critical model parameters are estimated. We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola Bay (southeastern United States) and provide a quantitative assessment of how the prior distributions affect the model results. The approach is also bench-marked with more conventional methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration). The EBASE approach is available using open source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

@@

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecosystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011]. Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005]. However, gross primary production and ecosystem respiration estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, changing spectral light quality with depth, the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are further compounded in estuaries, where high temporal and spatial variability have been noted in global syntheses of photosynthesis and respiration. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year to year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole-system respiration rates based on the open-water method [@hoellein2013].

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS). Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.

Another limitation of the open-water method is the estimation of the air-water oxygen flux, which is an important term in the oxygen budget. Most approaches estimate this flux using an empirically derived gas transfer velocity that is dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18]. Particularly challenging is that the exchange velocity in estuaries also depends on water depth and tides [@Ho16] and possibly additional factors, such as fetch and the concentration of total suspended solids [@Borges11]. The importance of the air--water gas flux in the oxygen budget opens up the possibility that the oxygen data themselves may be used to determine the gas transfer velocity, an approach that has been applied in streams [e.g., @riley2013].

A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Bayesian techniques have only been explored in limited estuarine applications despite their potential to address the above [@ciavatta2008; @tassone2019]. Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016]. These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of parameters related to gas exchange, P, and R (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange. A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15]. BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations. As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
